---
title: "Final Project"
output: html_document
date: "2024-12-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(usmap)

load(url("https://github.com/nbiler/Data-and-Society-Final/raw/refs/heads/main/NSDUH%20dataset/NSDUH_2023.Rdata"))
drug_df <- puf2023_102124
```


```{r}
drug_df %>% 
  select(ILLFLAG) %>% 
  table()
```


### By sex
### Need to do more research but this graph shows that sex really doesn't effect if one uses drugs.
### Maybe I should look at it in last year/month

```{r}
drug_df %>% 
  select(ILLFLAG, IRSEX) %>% 
  mutate(ILLFLAG_LAB = case_when(ILLFLAG == 0 ~ "Never Used",
                                 ILLFLAG == 1 ~ "Ever Used"),
         IRSEX_LAB = case_when(IRSEX == 1 ~ "Male",
                                IRSEX == 2 ~ "Female")) %>% 
  group_by(IRSEX_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop.sex = prop.table(n)) %>% 
  ggplot(aes(x = IRSEX_LAB, y = prop.sex, fill = ILLFLAG_LAB)) +
  geom_col()
```

### By sex for college students aged 18-22 over last 12 months
```{r}
drug_df %>% 
  filter(ENRLCOLLST2 == 1|
           ENRLCOLLST2 == 2) %>%
  select(ILLYR, IRSEX) %>% 
  mutate(ILLYR_LAB = case_when(ILLYR == 0 ~ "Did not use in past year",
                                 ILLYR == 1 ~ "Used in past year"),
         IRSEX_LAB = case_when(IRSEX == 1 ~ "Male",
                                IRSEX == 2 ~ "Female")) %>% 
  group_by(IRSEX_LAB, ILLYR_LAB) %>% 
  tally() %>% 
  mutate(prop.sex = prop.table(n)) %>% 
  ggplot(aes(x = IRSEX_LAB, y = prop.sex, fill = ILLYR_LAB)) +
  geom_col()
```


### By age
```{r}
drug_df %>% 
  select(ILLYR, CATAG6) %>%
  mutate(CATAG6_LAB = case_when(CATAG6 == 1 ~ "12-17 Years Old",
                                CATAG6 == 2 ~ "18-25 Years Old",
                                CATAG6 == 3 ~ "26-34 Years Old",
                                CATAG6 == 4 ~ "35-49 Years Old",
                                CATAG6 == 5 ~ "50-64 Years Old",
                                CATAG6 == 6 ~ "65 or Older"),
         ILLYR_LAB = case_when(ILLYR == 0 ~ "Did not use in past year",
                                 ILLYR == 1 ~ "Used in past year")) %>% 
  group_by(CATAG6_LAB, ILLYR_LAB) %>% 
  tally() %>% 
  mutate(prop.age = prop.table(n)) %>% 
  ggplot(aes(x = CATAG6_LAB, y = prop.age, fill = ILLYR_LAB)) +
  geom_col()
```


### By race
```{r}
drug_df %>% 
  select(ILLFLAG, NEWRACE2) %>% 
  mutate(
    ILLFLAG_LAB = factor(case_when(
      ILLFLAG == 0 ~ "Never Used",
      ILLFLAG == 1 ~ "Ever Used"
    ), levels = c("Never Used", "Ever Used")),
    
    NEWRACE2_LAB = case_when(
      NEWRACE2 == 1 ~ "NonHisp White",
      NEWRACE2 == 2 ~ "NonHisp Black/Afr Am",
      NEWRACE2 == 3 ~ "NonHisp Native Am/AK Native",
      NEWRACE2 == 4 ~ "NonHisp Native HI/Other Pac Is",
      NEWRACE2 == 5 ~ "NonHisp Asian",
      NEWRACE2 == 6 ~ "NonHisp more than one race",
      NEWRACE2 == 7 ~ "Hispanic"
    )) %>% 
  group_by(NEWRACE2_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop_race = prop.table(n)) %>% 
  ggplot(aes(x = NEWRACE2_LAB, y = prop_race, fill = ILLFLAG_LAB)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  

```

### By race for college students aged 18-22 over last 12 months
```{r}
drug_df %>% 
  filter(ENRLCOLLST2 == 1|
           ENRLCOLLST2 == 2) %>% 
  select(ILLYR, NEWRACE2) %>% 
  mutate(
    ILLYR_LAB = case_when(ILLYR == 0 ~ "Did not use in past year",
                                 ILLYR == 1 ~ "Used in past year"),
    
    NEWRACE2_LAB = case_when(
      NEWRACE2 == 1 ~ "NonHisp White",
      NEWRACE2 == 2 ~ "NonHisp Black/Afr Am",
      NEWRACE2 == 3 ~ "NonHisp Native Am/AK Native",
      NEWRACE2 == 4 ~ "NonHisp Native HI/Other Pac Is",
      NEWRACE2 == 5 ~ "NonHisp Asian",
      NEWRACE2 == 6 ~ "NonHisp more than one race",
      NEWRACE2 == 7 ~ "Hispanic"
    )) %>% 
  group_by(NEWRACE2_LAB, ILLYR_LAB) %>% 
  tally() %>% 
  mutate(prop_race = prop.table(n)) %>% 
  ggplot(aes(x = NEWRACE2_LAB, y = prop_race, fill = ILLYR_LAB)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### By race for college students aged 18-22 over last 12 months for female
```{r}
drug_df %>% 
  filter(ENRLCOLLST2 == 1|
           ENRLCOLLST2 == 2) %>% 
  filter(IRSEX == 2) %>% 
  select(ILLYR, NEWRACE2) %>% 
  mutate(
    ILLYR_LAB = case_when(ILLYR == 0 ~ "Did not use in past year",
                                 ILLYR == 1 ~ "Used in past year"),
    
    NEWRACE2_LAB = case_when(
      NEWRACE2 == 1 ~ "NonHisp White",
      NEWRACE2 == 2 ~ "NonHisp Black/Afr Am",
      NEWRACE2 == 3 ~ "NonHisp Native Am/AK Native",
      NEWRACE2 == 4 ~ "NonHisp Native HI/Other Pac Is",
      NEWRACE2 == 5 ~ "NonHisp Asian",
      NEWRACE2 == 6 ~ "NonHisp more than one race",
      NEWRACE2 == 7 ~ "Hispanic"
    )) %>% 
  group_by(NEWRACE2_LAB, ILLYR_LAB) %>% 
  tally() %>% 
  mutate(prop_race = prop.table(n)) %>% 
  ggplot(aes(x = NEWRACE2_LAB, y = prop_race, fill = ILLYR_LAB)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### By race for college students aged 18-22 over last 12 months for male
```{r}
drug_df %>% 
  filter(ENRLCOLLST2 == 1|
           ENRLCOLLST2 == 2) %>% 
  filter(IRSEX == 1) %>% 
  select(ILLYR, NEWRACE2) %>% 
  mutate(
    ILLYR_LAB = case_when(ILLYR == 0 ~ "Did not use in past year",
                                 ILLYR == 1 ~ "Used in past year"),
    
    NEWRACE2_LAB = case_when(
      NEWRACE2 == 1 ~ "NonHisp White",
      NEWRACE2 == 2 ~ "NonHisp Black/Afr Am",
      NEWRACE2 == 3 ~ "NonHisp Native Am/AK Native",
      NEWRACE2 == 4 ~ "NonHisp Native HI/Other Pac Is",
      NEWRACE2 == 5 ~ "NonHisp Asian",
      NEWRACE2 == 6 ~ "NonHisp more than one race",
      NEWRACE2 == 7 ~ "Hispanic"
    )) %>% 
  group_by(NEWRACE2_LAB, ILLYR_LAB) %>% 
  tally() %>% 
  mutate(prop_race = prop.table(n)) %>% 
  ggplot(aes(x = NEWRACE2_LAB, y = prop_race, fill = ILLYR_LAB)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Personal Income
```{r}
drug_df %>% 
  select(ILLFLAG, IRPINC3) %>% 
  mutate(
  ILLFLAG_LAB = factor(case_when(
    ILLFLAG == 0 ~ "Never Used",
    ILLFLAG == 1 ~ "Ever Used"
  ), levels = c("Never Used", "Ever Used")),
  
  IRPINC3_LAB = factor(case_when(
    IRPINC3 == 1 ~ "Less than $10,000 (Including Loss)",
    IRPINC3 == 2 ~ "$10,000 - $19,999",
    IRPINC3 == 3 ~ "$20,000 - $29,999",
    IRPINC3 == 4 ~ "$30,000 - $39,999",
    IRPINC3 == 5 ~ "$40,000 - $49,999",
    IRPINC3 == 6 ~ "$50,000 - $74,999",
    IRPINC3 == 7 ~ "$75,000 or more"
  ), levels = c(
    "Less than $10,000 (Including Loss)",
    "$10,000 - $19,999",
    "$20,000 - $29,999",
    "$30,000 - $39,999",
    "$40,000 - $49,999",
    "$50,000 - $74,999",
    "$75,000 or more"
  ))
) %>% 
  group_by(IRPINC3_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop_inc = prop.table(n)) %>% 
  ggplot(aes(x = IRPINC3_LAB, y = prop_inc, fill = ILLFLAG_LAB)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Family Income
```{r}
drug_df %>% 
  select(ILLFLAG, IRFAMIN3) %>% 
  mutate(
  ILLFLAG_LAB = factor(case_when(
    ILLFLAG == 0 ~ "Never Used",
    ILLFLAG == 1 ~ "Ever Used"
  ), levels = c("Never Used", "Ever Used")),
  
  IRFAMIN3_LAB = factor(case_when(
    IRFAMIN3 == 1 ~ "Less than $10,000 (Including Loss)",
    IRFAMIN3 == 2 ~ "$10,000 - $19,999",
    IRFAMIN3 == 3 ~ "$20,000 - $29,999",
    IRFAMIN3 == 4 ~ "$30,000 - $39,999",
    IRFAMIN3 == 5 ~ "$40,000 - $49,999",
    IRFAMIN3 == 6 ~ "$50,000 - $74,999",
    IRFAMIN3 == 7 ~ "$75,000 or more"
  ), levels = c(
    "Less than $10,000 (Including Loss)",
    "$10,000 - $19,999",
    "$20,000 - $29,999",
    "$30,000 - $39,999",
    "$40,000 - $49,999",
    "$50,000 - $74,999",
    "$75,000 or more"
  ))
) %>% 
  group_by(IRFAMIN3_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop_inc = prop.table(n)) %>% 
  ggplot(aes(x = IRFAMIN3_LAB, y = prop_inc, fill = ILLFLAG_LAB)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### By highest level of education 24 and older
```{r}
drug_df %>% 
  select(AGE3, ILLFLAG, EDUHIGHCAT) %>% 
  filter(AGE3 >= 6) %>% 
  mutate(ILLFLAG_LAB = factor(case_when(
    ILLFLAG == 0 ~ "Never Used",
    ILLFLAG == 1 ~ "Ever Used"
  ), levels = c("Never Used", "Ever Used")),
  EDUHIGHCAT_LAB = factor(case_when(
    EDUHIGHCAT == 1 ~ "Less high school",
    EDUHIGHCAT == 2 ~ "High school grad",
    EDUHIGHCAT == 3 ~ "Some coll/Assoc Dg",
    EDUHIGHCAT == 4 ~ "College graduate"),
    levels = c("Less high school", "High school grad",
               "Some coll/Assoc Dg", "College graduate")
  )) %>% 
  group_by(EDUHIGHCAT_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop_EDUC = prop.table(n)) %>% 
  ggplot(aes(x = EDUHIGHCAT_LAB, y = prop_EDUC, fill = ILLFLAG_LAB)) +
  geom_col()
  
```

### By sexuality
```{r}
drug_df %>% 
  select(ILLFLAG, SEXIDENT22) %>% 
  filter(SEXIDENT22 <= 5) %>% 
  mutate(ILLFLAG_LAB = factor(case_when(
    ILLFLAG == 0 ~ "Never Used",
    ILLFLAG == 1 ~ "Ever Used"
  ), levels = c("Never Used", "Ever Used")),
  SEXIDENT22_LAB = case_when(SEXIDENT22 == 1 ~ "Heterosexual",
                             SEXIDENT22 == 2 ~ "LGBTQ+",
                             SEXIDENT22 == 3 ~ "LGBTQ+",
                             SEXIDENT22 == 4 ~ "LGBTQ+",
                             SEXIDENT22 == 5 ~ "LGBTQ+")) %>% 
  group_by(SEXIDENT22_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop_sex = prop.table(n)) %>% 
  ggplot(aes(x = SEXIDENT22_LAB, y = prop_sex, fill = ILLFLAG_LAB)) +
  geom_col()
```

### If served in US armed forces
```{r}
drug_df %>% 
  select(ILLFLAG, SERVICE) %>% 
  mutate(ILLFLAG_LAB = factor(case_when(
    ILLFLAG == 0 ~ "Never Used",
    ILLFLAG == 1 ~ "Ever Used"
  ), levels = c("Never Used", "Ever Used")),
  SERVICE_LAB = case_when(SERVICE == 1 ~ "Yes",
                          SERVICE == 2 ~ "No")) %>% 
    filter(complete.cases(.)) %>% 
  group_by(SERVICE_LAB, ILLFLAG_LAB) %>% 
  tally() %>% 
  mutate(prop_SERVICE = prop.table(n)) %>% 
  ggplot(aes(x = SERVICE_LAB, y = prop_SERVICE, fill = ILLFLAG_LAB)) +
  geom_col()
```

### Mental health in past month vs used drugs in last month
```{r}
drug_df %>% 
  select(KSSLR6MON, ILLMON) %>%
  mutate(ILLMON_LAB = factor(case_when(
    ILLMON == 0 ~ "Did not use in past month",
    ILLMON == 1 ~ "Used in past month"
  ), levels = c("Did not use in past month", "Used in past month"))) %>% 
  group_by(KSSLR6MON, ILLMON_LAB) %>% 
  tally() %>% 
  mutate(prop_KSS = prop.table(n)) %>% 
  ggplot(aes(x = KSSLR6MON, y = prop_KSS, fill = ILLMON_LAB)) +
  geom_col()
```

### Employment over 18

```{r}
drug_df %>% 
  select(IRWRKSTAT18, ILLMON) %>% 
   mutate(ILLMON_LAB = factor(case_when(
    ILLMON == 0 ~ "Did not use in past month",
    ILLMON == 1 ~ "Used in past month"
  ), levels = c("Did not use in past month", "Used in past month")),
  IRWRKSTAT18_LAB = case_when(
    IRWRKSTAT18 == 1 ~ "Employed",
    IRWRKSTAT18 == 2 ~ "Employed",
    IRWRKSTAT18 == 3 ~ "Unemployed")) %>% 
  group_by(IRWRKSTAT18_LAB, ILLMON_LAB) %>% 
  tally() %>% 
  mutate(prop_work = prop.table(n)) %>% 
  ggplot(aes(x = IRWRKSTAT18_LAB, y = prop_work, fill = ILLMON_LAB)) +
  geom_col()
```

